---
title: "Sablefish escape ring study"
author: "Jane Sullivan, Andrew Olson, Aaron Baldwin, Ben Williams" 
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

YEAR <- 2019 # study year(s)
source("../code/helper.r")

bio <- read_csv(paste0("../data/bio_cleaned_", YEAR, ".csv")) %>% 
  filter(!is.na(length)) %>% 
  mutate(Treatment = derivedFactor("Control" = Treatment == "99",
                                   "3.50 in" = Treatment == "02",
                                   "3.75 in" = Treatment == "01",
                                   "4.00 in" = Treatment == "00",
                                   .default = NA,
                                   .ordered = TRUE))

# Effort data 
effort <- read_csv(paste0("../data/pot_effort_", YEAR, ".csv"))


```

```{r data_summary}

# Total lengths
tot_fish <- bio %>% 
  dplyr::summarise(n = n()) %>% 
  pull()

# Samples that don't have necessary treatment data associated with them. See GitHub issue #1
discards <- bio %>% 
  filter(is.na(Treatment)) %>% 
  group_by(discard_status) %>% 
  dplyr::summarise(n = n()) %>% 
  dplyr::rename(`Discard status` = discard_status) 

tot_discards <- bio %>% 
  filter(is.na(Treatment)) %>% 
  dplyr::summarise(n = n()) %>% 
  mutate(`Discard status` = "Total")

discards_tbl <- bind_rows(discards, tot_discards) %>% 
  mutate(n = prettyNum(n, big.mark = ","))

# Remove these fish from analysis
bio <- bio %>% filter(!is.na(Treatment))

# Final sample sizes for fork lengths (cm)
len_n <- bio %>% 
  group_by(Treatment) %>% 
  dplyr::summarise(len_n = n()) 

len_totn <- bio %>% 
              dplyr::summarise(len_n = n()) %>% 
              mutate(Treatment = "Total")

len_n <- len_n %>% 
  bind_rows(len_totn)

# Girths (mm)
girths_n <- bio %>% 
  filter(!is.na(girth)) %>% 
  group_by(Treatment) %>% 
  dplyr::summarise(girth_n = n()) %>% 
  bind_rows(bio %>% 
              filter(!is.na(girth)) %>% 
              dplyr::summarise(girth_n = n()) %>%
              mutate(Treatment = "Total"))

# Weights (mm)
wt_n <- bio %>% 
  filter(!is.na(weight)) %>% 
  group_by(Treatment) %>% 
  dplyr::summarise(wt_n = n()) %>% 
  bind_rows(bio %>% 
              filter(!is.na(weight)) %>% 
              dplyr::summarise(wt_n = n()) %>%
              mutate(Treatment = "Total"))

# Stuck in rings
stuck_n <- bio %>% 
  filter(grepl(c("Stuck|stuck"), comments)) %>% 
  group_by(Treatment) %>% 
  dplyr::summarise(stuck_n = n()) %>% 
  bind_rows(bio %>% 
              filter(grepl(c("Stuck|stuck"), comments)) %>% 
              dplyr::summarise(stuck_n = n()) %>%
              mutate(Treatment = "Total"))

# Final sample size table  
n_tbl <- left_join(len_n, girths_n) %>% 
  left_join(wt_n) %>% 
  left_join(stuck_n)

# Format
n_tbl[c(2,3,4,5)] <- lapply(n_tbl[c(2,3,4,5)], prettyNum, big.mark=",")
names(n_tbl) <- c("Treatment", "Fork length", "Girth", "Weight", "Stuck in escape ring")
kable(n_tbl)
```

There were `r prettyNum(pull(tot_discards, n), big.mark = ',')` fish that were lost, not tagged properly,  already tagged from another agency, dead or unmarketable due to sandflea bites or other damage, and were subsequently excluded from the remainder of the analysis. This reflected `r formatC(round(pull(tot_discards, n) / tot_fish * 100, 2))` of the total fish captured on the survey (`r prettyNum(tot_fish, big.mark = ",")`), resulting in a total of `r prettyNum(pull(len_totn, len_n), big.mark = ",")` fish lengths available for analysis (T). In addition to lengths, a subset of sablefish girths and weights were collected using a stratified random sampling design during the tagging survey. The fish girth, or circumference, was measured anterior to the dorsal fin and posterior to the pectoral and pelvic fins at fish's the thickest point and recorded to the nearest mm. Additionally, fish stuck in the esacape rings were documented and measured for girth and weight when possible.


